The hydra and the rooster

### 2020-07-23 latex/chapter-hydras.tex
redaction introduction to ordinals

\subsection{Big steps}

@@

In this section we show how long a simple hydra battle can last.  The following example
considers a very small hydra, shown on figure~\vref{fig:hinit}.

@@

Moreover, in order to \emph{compute} the length of a battle, we set Hercules strategy : to chop the rightmost head at each round, and the hydra's strategy : at the $i$-th round, the replication factor is exactly $i$ (which means that the battle is
a \emph{standard} battle, see section~\vref{sect:battle-classes}).

The  battle is so long that no \emph{test} can give us an estimation of its length, and we do need the expressive power of logic to compute this length. However, in order to  guess this length, we made some experiments, computing with \gallina{}, \coq{}'s  functional programming language.
Thus, we can consider this development as a collaboration of proof with computation.

We want to verify whether Hercules wins a standard battle starting from \texttt{hinit} at time $t=0$ , and in this case, in how many rounds. We assume Hercules' 
strategy is to chop off the rightmost head at every round.

@@

During the two first rounds, our hydra loses its two rightmost heads. Thus just before the third round, it looks like in figure~\vref{fig:hinit-plus2}.

@@

The following lemma  is a formal statement about the two first rounds of the battle.

@@

@@

We would like to have a formal proof that the length of the considered battle is 
equal to some integer $N$, but we have first  to \emph{guess} this number.
In order to study \emph{experimentally} the different  configurations of the  battle, we will use a simple datatype for representing the states as tuples composed of
the round number, and the respective number of daughters  \texttt{h2}, \texttt{h1}, and heads
of the current hydra.

@@

We are now able to reason about the sequence of transitions defined by our hydra battle. Instead of using the data-type \texttt{state} we study the relationship
between different configurations of the battle.

@@

We are now able to make giant steps in the simulation of the battle.
First, we iterate the lemma \texttt{reachable\_S}.

@@

@@

Let $b$ be an instance of class \texttt{Battle}. A \emph{variant} for $b$ consists
in a well-founded relation $<$  on some type \texttt{A}, and a function
(also called a \emph{measure}) \texttt{m:Hydra->A} such that for any successive steps $(i,h)$ and $(1+i,h')$  of a battle in $b$, the inequality $m(h')<m(h)$ holds.

@@

\begin{Coqsrc}
Class Hvariant {A:Type}{Lt:relation A}(Wf: well_founded Lt)(b : Battle)
                          (m: Hydra -> A): Prop :=
 {variant_decr :forall i h h', 
     h <> head -> b i  h h' -> Lt (m h') (m h)}.
\end{Coqsrc}

@@

\begin{exercise}
 Prove that, if there is an instance of \texttt{Hvariant Lt wf\_Lt b m}, then there exists no infinite battle in  \texttt{b}.
\end{exercise}

@@

The class \texttt{Hvariant} is parameterized by a well-founded order $<$ on some type $A$, a class of battles $B$ and a measure $m$. When $<$ or $m$ are 
badly given for a given class $B$, one may fail to find a variant.

We show that this is the case of the type \texttt{nat} of Peano natural numbers, provided with the usual strict order $<$.

\subsubsection{Using the Hydra's height as a variant}
Trivially, a measure like the height will not work for building a variant.

Please consider the hydras of Fig.~\vref{fig:h-omega2-small} and
\ref{fig:h-omega-omega}. The former can be transformed into the latter in one round, but their heights are equal.

\begin{figure}[htb]
\centering
\begin{tikzpicture}[very thick, scale=0.5]
\node (foot) at (2,0) {$\bullet$};
\node (N1) at (2,2) {$\bullet$};
\node (N2) at (2,4) {$\Smiley[2][green]$};
\draw (foot) -- (N1);
\draw (N1) -- (N2) ;
\end{tikzpicture}
\caption{\label{fig:h-omega-omega}}
\end{figure}

\vspace{4pt}
\emph{From Module~\href{../src/html/hydras.Hydra.Omega_Small.html\#height_bad}{Hydra.Hydra.Omega\_Small}}

\begin{Coqsrc}
Lemma height_bad :  ~ Hvariant lt_wf free height.
Proof.
  intros [H];
   specialize (H 1 (hyd1 (hyd2 head head))  (hyd1 (hyd1 head)));
   apply (lt_irrefl 2), H.
  -  apply reachable_free.
  -  discriminate.
  -  exists 0; right;  R2_here 0;  left. 
Qed.
\end{Coqsrc}

\subsubsection{Using any variant defined on \texorpdfstring{nat}{nat}}
\label{sect:omega-case}
We could imagine that our previous failure is due to the choice of \texttt{height} as a measure, and that a more complex function would work.
In fact, we can prove that 
\emph{no} instance of class \texttt{WfVariant round Peano.lt $m$} can be build, where
$m$ is \emph{any} function of type \texttt{Hydra $\arrow$ nat}.

@@

In order to build a variant for proving the termination of all hydra battles, we need to consider order structures more complex than the usual order on type \texttt{nat}. The next chapter will present the notion of \emph{ordinal notation} from the point view of the \coq{} user: inductive types, functions, type classes, etc. 
The ordinal $\epsilon_0$, the first one allowing us to build our desired variant, will be studied in detail in 
Chapter~\ref{chap:T1}.

>>>>

\subsection{Formal definition of battles}

@@

In this section we consider a simple example of battle, starting with a small hydra,
shown on figure~\vref{fig:hinit}, with a simple strategy for both players:

\begin{itemize}
\item Hercules chops always the rightmost head of the hydra.
\item The battle is standard: at the round number $i$, the expected replication is $i$.
\end{itemize}

@@

The lemma we would like to prove is ``The considered battle lasts exactly $N$ rounds'',
with $N$ being a natural number we gave to guess.

But the  battle is so long that no \emph{test} can give us an estimation of its length, and we do need the expressive power of logic to compute this length. However, in order to  guess this length, we made some experiments, computing with \gallina{}, \coq{}'s  functional programming language.
Thus, we can consider this development as a collaboration of proof with computation.

In the following lines, we try to show faithfully how we found the value of the number $N$.

@@

\subsection{The beginning of hostilities}
During the two first rounds, our hydra loses its two rightmost heads. Thus just before the third round, it looks like in figure~\vref{fig:hinit-plus2}.

@@

The following lemma  is a formal statement about these first rounds, in terms of the
\texttt{battle} predicate.

@@

\subsection{Looking for regularities}

@@

\subsection{Computing \dots}
In order to study \emph{experimentally} the different  configurations of the  battle, we will use a simple datatype for representing the states as tuples composed of
the round number, and the respective number of daughters  \texttt{h2}, \texttt{h1}, and heads
of the current hydra.

@@

\subsection{Proving \dots}
We are now able to reason about the sequence of transitions defined by our hydra battle. Instead of using the data-type \texttt{state} we study the relationship
between different configurations of the battle.

@@

\subsection{Giant steps}

We are now able to make bigger steps in the simulation of the battle.
First, we iterate the lemma \texttt{reachable\_S}.

@@

\subsection{A minoration lemma}

@@

Let $B$ be an instance of class \texttt{Battle}. A \emph{variant} for $B$ consists
in a well-founded relation $<$  on some type \texttt{A}, and a function
(also called a \emph{measure}) \texttt{m:Hydra->A} such that for any successive steps $(i,h)$ and $(1+i,h')$  of a battle in $B$, the inequality $m(h')<m(h)$ holds.

@@

\begin{Coqsrc}
Class Hvariant {A:Type}{Lt:relation A}(Wf: well_founded Lt)(B : Battle)
  (m: Hydra -> A):   Prop :=
  {variant_decr :forall i h h',
      h <> head ->
      battle_r  B i  h h' -> Lt (m h') (m h)}.
\end{Coqsrc}

@@

\begin{exercise}
 Prove that, if there is an instance of \texttt{Hvariant Lt wf\_Lt $B$ $m$}, then there exists no infinite battle in  $B$.
\end{exercise}

@@

The best known well-founded order is the natural order on the set $\mathbb{N}$ of natural numbers (the type \texttt{nat} of Standard library). It would be interesting to look for some measure $m:\texttt{nat}\arrow\texttt{nat}$ and prove it is a variant.

Unfortunately, we can prove that 
\emph{no} instance of class \texttt{WfVariant round Peano.lt $m$} can be built, where
$m$ is \emph{any} function of type \texttt{Hydra $\arrow$ nat}.

@@

In order to build a variant for proving the termination of all hydra battles, we need to consider order structures more complex than the usual order on type \texttt{nat}. 
The notion of \emph{ordinal number} provides a catalogue of well-founded order types.
For a reasonably large bunch of ordinal numbers, \emph{ordinal notations} are data-types which allow the \coq{} user to define functions, to compute and prove some properties, for instance by reflection.

So, the next chapters will present ordinal numbers and ordinal notations, and we will
look for a variant complex enough for proving the termination of all battles, but not too complex if possible.

### 2020-07-23 latex/chapter-ordinals.tex
redaction introduction to ordinals

\chapter{Introduction to ordinal numbers with the Coq proof assistant}

\centerline{\LARGE {\bf work in progress!}}

@@

>>>>

\chapter{Introduction to ordinal numbers}

@@

More intuitively, let us have a look at Figure~\ref{fig:ordinal-sequence}. It presents a sequence of ordinal numbers, as an extension of the sequence of natural numbers.

\begin{figure}[h]
  \centering
\fbox{\Large
  \begin{minipage}{1.0\linewidth}
  \begin{align*}
     &\textcolor{blue}{0},\,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,\ldots\\
&\textcolor{red}{\omega},\,\omega+1,\omega+2,\omega+3,\ldots\\
&\textcolor{red}{\omega\times 2},\,\omega\times 2+1,\ldots, \textcolor{red}{\omega\times 3},\,\omega\times 3+1,\ldots, \textcolor{red}{\omega\times 4},\ldots,\\
&\textcolor{red}{\omega^2},\ldots, \textcolor{red}{\omega^2\times 42},\ldots,\textcolor{red}{\omega^3},\ldots, \textcolor{red}{\omega^4},\omega^4+1,\ldots,\\
&\textcolor{red}{\omega^\omega2},\ldots, \textcolor{red}{\omega^\omega\times 2},\ldots,\\
&\textcolor{red}{\omega^{\omega^\omega}},\ldots, \textcolor{red}{\omega^{\omega^\omega}+\omega^\omega\times 42+ \omega^{55}+\omega}, \ldots, \textcolor{red}{\omega^{\omega^{\omega+1}}}, \omega^{\omega^{\omega+1}}+1,\dots\\
& \textcolor{red}{\epsilon_0 (= \omega^\epsilon_0)}, \epsilon_0+1, \ldots,\\
& \textcolor{red}{\epsilon_1}, \ldots, \textcolor{red}{\epsilon_2}, \ldots, \textcolor{red}{\epsilon_\omega},\ldots \\
& \textcolor{red}{\Gamma_0}, \Gamma_0+1, \Gamma_0+2, \Gamma_0+3,\ldots, \textcolor{red}{\Gamma_0+\omega}, \ldots,\\
&\ldots
  \end{align*}   
  \end{minipage}}

  \caption{A short overview of the sequence of ordinal numbers}
  \label{fig:ordinal-sequence}
\end{figure}

Let us comment some features of this table.

\begin{itemize}
\item The ordinals are listed in a strictly increasing order. 
\item Dots : ``$\ldots$'' stand for an infinite sequence of ordinals, not shown for lack of space. For instance, the ordinal $42$ is not shown in the first line, but it exists, betaween $41$ and $43$.
\item Each ordinal in black is an immediate successor of anotter ordinal. We call it a \emph{successor} ordinal. For instance, $43$ is the successor of $42$, and $\omega^4+1$ the successor of $\omega^4$.
\item Ordinals that follow dots (written in red) are called \emph{limit ordinals}. With respect to the order induced by this sequence, any limit ordinal $\alpha$ is the least upper bound of  the set $\mathbb{O}_\alpha$ of all ordinals strictly less than $\alpha$.
For instance $\omega$ is the least upper bound of the set of all finite ordinals (in the first line). $\omega$ is also the first limit ordinal, and the first infinite ordinal, in the sense that 
the set $\mathbb{O}_\omega$ is infinite.
\item The ordinal $\epsilon_0$ is the first number which is equal to its own exponential of base $\omega$. It plays an important role in proof theory, and is particularly studied in chapters~\ref{chap:T1} to \ref{chap:alpha-large}.
\end{itemize}
